Possible Inconsistency in Maximum Likelihood Calculation of Ancestral States

### 2017-03-21 main.tex
rearranging

To simplify matters, we will first focus on bounding the the terms with the generating probabilities as they are more complicated.
Assume $\tau^*=\tau_1$ and $t^*=\{x,y,y\}$ so that
\begin{align}
\label{eq:information}
H_{\tau'}(x, y;x', y', w') &= \sum_{j} P(s_j | \tau_1, \{x,y,y\}) \cdot \log P(s_j | \tau', \{x',y',w'\}) \nonumber \\ 
&:= \sum_{j} p(s_j) \cdot \log P(s_j | \tau', \{x',y',w'\}).
\end{align}
By Gibbs inequality,
$$
H_{\tau'}(x, y;x', y', w') \le H_{\tau'}(x, y;x, y, y),
$$
though this bound can be made tighter.

\textbf{Show}: where these simplifications come about.
It's just adding/subtracting $\log$ terms, but could be good to walk through.

In our simplifications we will make use of the constants
$$
a = p(0100) + p(1000) + p(0011) + p(1001) = (4-2x^2-2y^2)/8,
$$
$$
b = p(1000) + p(0010) + p(0011) + p(1001) = (4-2x^2-2y^2)/8,
$$
$$
c = p(0001) + p(0010) + p(0100) + p(1000) = (4-4x^2y^2)/8
$$
and
$$
d = p(0101) + p(1001) = (2-4xy^2+2x^2y^2)/8
$$
with corresponding bounds
$$
0 \le a,b,c \le \frac{1}{2}
$$
and
$$
0 \le d \le \frac{1}{4}.
$$

Our key approach will be to decompose the likelihood as
\begin{equation}
\label{eq:partial-likelihood}
L^{(i)}(x', y', w') = H_{\tau'}(x,y;x',y',w') + L^{p,(i)}(x',y',w')
\end{equation}
where the first term is \eqref{eq:information} and the second term will be a weighted sum of the $\log$s of values from Table~\ref{tab:likelihoods}, the superscript indicating which choices for ancestral state likelihood values we choose.
Define
$$
\{\hat{x}, \hat{y}, \hat{w}, \hat{i}\} = \arg\max_{\{x,y,w,i\}} \ L^{p,(i)}(x,y,w).
$$
We have upper and lower bounds for \eqref{eq:partial-likelihood} in the form
$$
H_{\tau'}(x,y; \hat{x}, \hat{y}, \hat{w}) + L^{p,(\hat{i})}(\hat{x}, \hat{y}, \hat{w}) \le \max_{x,y,w} \ L^{(i)}(x,y,w) \le H_{\tau'}(x,y;x,y,y) + L^{p,(\hat{i})}(\hat{x}, \hat{y}, \hat{w}).
$$

@@

\subsubsection{A simplified Farris likelihood}

The likelihood for the Farris topology is less complicated than that of the Felsenstein since only one site pattern has an ambiguous likelihood---all site patterns except \texttt{0101} have a fixed value for their likelihood when taking the maximum.
To fix this value, we will consider an upper bound on the partial likelihood by setting the value for site pattern 0011 to $(1+x)^2(1+w)(1+y)^2$.
Here the partial likelihood $L_{fa}^{p,(i)}(x,y,w)$ takes the form:
$$
L_{fa}^{p} = (2-a)\log(1+x)+a\log(1-x)+(2-b)\log(1+y)+b\log(1-y)+\log(1+w),
$$
and this has a maximum
$$
\hat{L}_{fa}^{p} = (2-a)\log(2-a)+a\log(a)+(2-b)\log(2-b)+b\log(b)+\log(2).
$$

\subsubsection{A simplified Felsenstein likelihood}

We construct similar bounds on the Felsenstein likelihood, though to make matters simpler we replace all $(1+y)$ terms in Table~\ref{tab:likelihoods} with $(1-y)$ to obtain a lower bound.
Unlike the Farris partial likelihood, the likelihood in this case has one of six forms:
$$
L_{fe}^{(1)} = 2\log(1+x)+(2-c-2d)\log(1+y)+(c+2d)\log(1-y)+\log(1-w),
$$
$$
L_{fe}^{(2)} = (2-c-2d)\log(1+x)+(c+2d)\log(1-x)+2\log(1+y)+\log(1-w),
$$
$$
L_{fe}^{(3)} = (2-2d)\log(1+x)+2d\log(1-x)+(2-c)\log(1+y)+c\log(1-y)+\log(1-w),
$$
$$
L_{fe}^{(4)} = (2-c)\log(1+x)+c\log(1-x)+(2-2d)\log(1+y)+2d\log(1-y)+\log(1-w),
$$
$$
L_{fe}^{(5)} = (2-d)\log(1+x)+d\log(1-x)+(2-c-d)\log(1+y)+(c+d)\log(1-y)+\log(1-w)
$$
and
$$
L_{fe}^{(6)} = (2-c-d)\log(1+x)+(c+d)\log(1-x)+(2-d)\log(1+y)+d\log(1-y)+\log(1-w)
$$
with maxima
$$
\hat{L}_{fe}^{(1)} = \hat{L}_{fe}^{(2)} = 2\log(2)+(2-c-2d)\log(2-c-2d)+(c+2d)\log(c+2d)
$$
$$
\hat{L}_{fe}^{(3)} = \hat{L}_{fe}^{(4)} = (2-2d)\log(2-2d)+2d\log(2d)+(2-c)\log(2-c)+c\log(c)
$$
and
$$
\hat{L}_{fe}^{(5)} = \hat{L}_{fe}^{(6)} = (2-d)\log(2-d)+d\log(d)+(2-c-d)\log(2-c-d)+(c+d)\log(c+d).
$$

@@

\subsubsection{Example}

Set $\{x, y\} = \{0.25, 0.25\}$.
For a Farris zone generating topology, we have
$$
a = b = 0.46875
$$
resulting in
$$
\hat{L}_{fa}^{p}(\hat{x}, \hat{y}, \hat{w}) \approx -2.17803560694
$$
and
$$
H_{fa}(x,y;x,y,y) \approx -2.07509999739.
$$
Computing the upper bound yields
$$
\max_{x,y,w} \ L_{fa}(x,y,w) \le -4.25313560433
$$
and it remains to show that 
$$
\max_{x,y,w} \ L_{fe}(x,y,w) \ge -4.25313560433
$$

For the same generating parameters,
$$
c = 0.498046875, \ d = 0.2431640625,
$$
resulting in
$$
\hat{L}_{fe}(\hat{x}, \hat{y}, \hat{w}) \approx -2.07919739112
$$
Calculating shows%
    %
    \footnote{It's not a typo that $\hat{L}_{fe}$ and $H_{fe}(x,y; \hat{x}, \hat{y}, \hat{w})$ are the same---at least not in the code.
        I should try to show why this is the case.
        }
    %
$$
H_{fe}(x,y; \hat{x}, \hat{y}, \hat{w}) \approx -2.07919739112
$$
and
$$
\max_{x,y,w} \ L_{fe}(x,y,w) \ge -4.15839478224 > -4.25313560433 \ge \max_{x,y,w} \ L_{fa}(x,y,w)
$$
where site patterns are generated from the Farris zone topology.
Confirming, run \texttt{python}'s numerical optimization solver for $\{.25, .25\}$.
Since we are computing a constrained maximization, we can find a global maximum using basin hopping, and doing so yields
$$
\max_{x,y,w} \ L_{fa}(x,y,w) \approx -4.5101771294218045
$$
and
$$
\max_{x,y,w} \ L_{fe}(x,y,w) \approx -4.1544859851260281.
$$
Woohoo!

\subsubsection{General example}

@@

>>>>

Assume $\tau^*=\tau_1$ and $t^*=\{x,y,y\}$.
Our approach will be to decompose the likelihood in \eqref{eq:site_pattern_profile_likelihood_mean} as
\begin{equation}
\label{eq:partial-likelihood}
L^{(i)}(x', y', w') = H(x,y,\tau_1;x',y',w', \tau') + L^{p,(i)}(x',y',w')
\end{equation}
where the first term is
\begin{align}
\label{eq:information}
H(x,y,\tau_1;x',y',w',\tau') &= \sum_{j} P(s_j | \tau_1, \{x,y,y\}) \cdot \log P(s_j | \tau', \{x',y',w'\}) \nonumber \\ 
&:= \sum_{j} p(s_j) \cdot \log P(s_j | \tau', \{x',y',w'\}).
\end{align}
and the second term will be a weighted sum of the $\log$s of values from Table~\ref{tab:likelihoods}, the superscript indicating which choices for ancestral state likelihood values we choose.

To bound the first term, we can make use of Gibbs inequality
$$
H(x, y, \tau_1;x', y', w', \tau') \le H(x, y, \tau_1;x, y, y, \tau_1),
$$
though this bound can be made tighter.
For the second term, we seek an upper bound for the $\tau_1$ case and a lower bound for the $\tau_2$ case.
Define
$$
\{\hat{x}, \hat{y}, \hat{w}, \hat{i}\} = \arg\max_{\{x,y,w,i\}} \ L^{p,(i)}(x,y,w,\tau').
$$
We have upper and lower bounds for \eqref{eq:partial-likelihood} in the form
$$
H(x,y,\tau_1; \hat{x}, \hat{y}, \hat{w}, \tau') + L^{p,(\hat{i})}(\hat{x}, \hat{y}, \hat{w},\tau_1) \le \max_{x,y,w} \ L^{(i)}(x,y,w) \le H(x,y,\tau_1;x,y,y,\tau_1) + L^{p,(\hat{i})}(\hat{x}, \hat{y}, \hat{w},\tau_1).
$$

@@

\subsubsection{Upper bound for Farris likelihood}

To obtain an upper bound for the Farris likelihood---the terms in the top half of Table~\ref{tab:likelihoods}---we use the term $(1+x)^2(1+w)(1+y)^2$ for the $0101$ site pattern.
Since all cases have a $(1+x)^2$ term except $\{0100, 1000, 0011, 1001\}$, and all cases except $\{0001, 0010, 0011, 1001\}$ have a $(1+y)^2$, we can add and subtract the $\log$ of these terms weighted by each of their generating probabilities.
In other words, define
$$
a = p(0100) + p(1000) + p(0011) + p(1001)
$$
and
$$
b = p(1000) + p(0010) + p(0011) + p(1001).
$$
Then $L^{p,(i)}(x',y',w',\tau_1)$ takes the form
$$
L^{p}(x',y',w') = (2-a)\log(1+x')+a\log(1-x')+(2-b)\log(1+y')+b\log(1-y')+\log(1+w')
$$
with a maximum
$$
\hat{L}^{p}(a,b,\tau_1) = (2-a)\log(2-a)+a\log(a)+(2-b)\log(2-b)+b\log(b)+\log(2)
$$
where $a$ and $b$ are functions of the true generating branch parameters $\{x,y\}$.

\subsubsection{Lower bound for Felsenstein likelihood}

We construct a similar lower bound on the Felsenstein likelihood, though here we replace all $(1+w)$ terms with $(1-w)$.
Let
$$
c = p(0001) + p(0010) + p(0100) + p(1000)
$$
and
$$
d = p(0101) + p(1001).
$$
The likelihood in this case takes one of six forms:
$$
L^{(1)}(x',y',w') = 2\log(1+x')+(2-c-2d)\log(1+y')+(c+2d)\log(1-y')+\log(1-w'),
$$
$$
L^{(2)}(x',y',w') = (2-c-2d)\log(1+x')+(c+2d)\log(1-x')+2\log(1+y')+\log(1-w'),
$$
$$
L^{(3)}(x',y',w') = (2-2d)\log(1+x')+2d\log(1-x')+(2-c)\log(1+y')+c\log(1-y')+\log(1-w'),
$$
$$
L^{(4)}(x',y',w') = (2-c)\log(1+x')+c\log(1-x')+(2-2d)\log(1+y')+2d\log(1-y')+\log(1-w'),
$$
$$
L^{(5)}(x',y',w') = (2-d)\log(1+x')+d\log(1-x')+(2-c-d)\log(1+y')+(c+d)\log(1-y')+\log(1-w')
$$
or
$$
L^{(6)}(x',y',w') = (2-c-d)\log(1+x')+(c+d)\log(1-x')+(2-d)\log(1+y')+d\log(1-y')+\log(1-w').
$$
These have maxima
$$
\hat{L}^{(1)}(c,d) = \hat{L}^{(2)}(c,d) = 2\log(2)+(2-c-2d)\log(2-c-2d)+(c+2d)\log(c+2d),
$$
$$
\hat{L}^{(3)}(c,d) = \hat{L}^{(4)}(c,d) = (2-2d)\log(2-2d)+2d\log(2d)+(2-c)\log(2-c)+c\log(c)
$$
and
$$
\hat{L}^{(5)}(c,d) = \hat{L}^{(6)}(c,d) = (2-d)\log(2-d)+d\log(d)+(2-c-d)\log(2-c-d)+(c+d)\log(c+d)
$$
where $c$ and $d$ are also functions of the true generating branch parameters $\{x,y\}$.

@@

Thus, let
$$
\hat{L}^{p}(c,d,\tau_2) = 2\log(2)+(2-c-2d)\log(2-c-2d)+(c+2d)\log(c+2d).
$$

\subsubsection{Region of inconsistency}

@@

\paragraph{Example}

Set $\{x, y\} = \{0.25, 0.25\}$.
For a Farris zone generating topology, we have
$$
a = b = 0.46875
$$
resulting in
$$
\hat{L}^{p}(\hat{x}, \hat{y}, \hat{w},\tau_1) \approx -2.17803560694
$$
and
$$
H(x,y,\tau_1;x,y,y,\tau_1) \approx -2.07509999739.
$$
Computing the upper bound yields
$$
\max_{x,y,w} \ L(x,y,w,\tau_1) \le -4.25313560433
$$
and it remains to show that 
$$
\max_{x,y,w} \ L(x,y,w,\tau_2) \ge -4.25313560433
$$

For the same generating parameters,
$$
c = 0.498046875, \ d = 0.2431640625,
$$
resulting in
$$
\hat{L}(\hat{x}, \hat{y}, \hat{w},\tau_2) \approx -2.07919739112
$$
Calculating shows%
    %
    \footnote{It's not a typo that $\hat{L}_{fe}$ and $H_{fe}$ are the same---at least not in the code.
        I should try to show why this is the case.
        }
    %
$$
H(x,y,\tau_1; \hat{x}, \hat{y}, \hat{w},\tau_2) \approx -2.07919739112
$$
and
$$
\max_{x,y,w} \ L(x,y,w,\tau_2) \ge -4.15839478224 > -4.25313560433 \ge \max_{x,y,w} \ L(x,y,w,\tau_1)
$$
where site patterns are generated from the Farris zone topology.
Confirming, run \texttt{python}'s numerical optimization solver for $\{.25, .25\}$.
Since we are computing a constrained maximization, we can find a global maximum using basin hopping, and doing so yields
$$
\max_{x,y,w} \ L(x,y,w,\tau_1) \approx -4.5101771294218045
$$
and
$$
\max_{x,y,w} \ L(x,y,w,\tau_2) \approx -4.1544859851260281.
$$
Woohoo!
